---
title: "Comparison of dN/dS ratios of Human TFs"
subtitle: "MyoD1: Myoblast determination protein 1"
output: 
  html_document: 
    theme: cosmo
    highlight: "pygments"
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: TRUE
---

In this markdown we are going to compare the dN/dS values of MyoD1 with the rest of the bHLH TFs. This way we can have an estimate of the selection pressure on the coding sequence of MyoD1.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/claudiavicentecomorera/Desktop/MSc/PGB/PGBProject")
```

# Import libraries

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(naniar)
library(ggplot2)
library(GO.db)
library(biomaRt)
library(ggpubr)
library(pheatmap)
```

# Read the data

```{r}
data <- read.table("HumanTFs_DBD.txt", header = T, row.names = 1, sep = "\t")
dnds <- read.table("HumanDnDsW.txt", header = T,  sep = "\t")

bHLH_data <- left_join(data[data$DBD == "bHLH",], dnds, by=c("Ensembl.ID"="GeneID"))
head(bHLH_data)
```

# Data inspection

```{r}
summary(bHLH_data)
vis_miss(bHLH_data)
#bHLH_data <- bHLH_data %>%
#  filter(!is.na(Homo_sapiens.dN) & !is.na(Homo_sapiens.dS) & !is.na(Homo_sapiens.w))
bHLH_data$Homo_sapiens.w[bHLH_data$HGNC.symbol == "MYOD1"]
```

```{r}
s <- bHLH_data[bHLH_data$HGNC.symbol %in% c("MYOD1", "NEUROG2", "NEUROD2", "ASCL1", "OLIG2", "PTF1A", "ATOH1"), c("Homo_sapiens.dN", "Homo_sapiens.dS", "Homo_sapiens.w"), drop = FALSE]
rownames(s) <- bHLH_data[bHLH_data$HGNC.symbol %in% c("MYOD1", "NEUROG2", "NEUROD2", "ASCL1", "OLIG2", "PTF1A", "ATOH1"), "HGNC.symbol"]
colnames(s) <- c("dN", "dS", "ω")
pheatmap(scale(s), cluster_rows = FALSE, cluster_cols = FALSE, color = colorRampPalette(c("blue", "white", "red"))(50), na_col = "#DDDDDD")
```


# Data log-transformation

```{r}
hist(bHLH_data$Homo_sapiens.w, main = "dN/dS Distribution", xlab = "dN/dS", breaks = 30, col = "#8EEE8B")
bHLH_data$Homo_sapiens.w.log2 <- log2(bHLH_data$Homo_sapiens.w)
hist(bHLH_data$Homo_sapiens.w.log2, main = "dN/dS Distribution", xlab = "dN/dS", breaks = 30, col = "#8EEE8B")
```

# Data visualization

MyoD1 dN/dS (ω) rato is 1e-04.

```{r warning=FALSE}
ggplot(bHLH_data, aes(x = HGNC.symbol, y = Homo_sapiens.w.log2)) +
  geom_bar(stat = "identity", aes(fill = HGNC.symbol == "MYOD1")) +  
  scale_fill_manual(values = c("#68B0F8", "#C968F8"), guide = "none") +  
  labs(title = "Bar plot of dN/dS (ω) values from bHLH family members", y = "dN/dS (ω)", x = "bHLH transcription factors") +
  theme_bw() + 
  scale_x_discrete(breaks = bHLH_data$HGNC.symbol[seq(1, nrow(bHLH_data), by = 2)], labels = function(x) {ifelse(x == "MYOD1", expression(bold("MYOD1")), x)}) +
  theme(axis.text.x = element_text(hjust = 1, size = 10), axis.text.y = element_text(size = 10))+
  rotate_x_text(angle = 90)
```

# dN/dS comparison using GO terms containing "muscle"

```{r}
xx <- as.list(GOTERM)
terms <- lapply(xx, Term)
muscle_terms <- grep("muscle", terms, ignore.case = TRUE)
muscle_terms <- unlist(terms[muscle_terms])

ensembl = useMart(biomart="ensembl", dataset="hsapiens_gene_ensembl")
muscle_gene <- getBM(attributes=c('ensembl_gene_id', "go_id"),
                     filters = 'go', values = names(muscle_terms), mart = ensembl)

df_muscle <- filter(bHLH_data, Ensembl.ID %in% unique(muscle_gene$ensembl_gene_id))

df_muscle$Geneset <- "Muscle"
bHLH_data$Geneset <- "Others"
df <- bHLH_data %>%
  rows_update(df_muscle, by = "Ensembl.ID")
head(df)
```

MyoD1, as the majority of proteins, seems to be under strong purifying selection. In the following violin plot bHLH members with biological functions related to muscle seem to have a lower ω rate compared to the members with other functions.

```{r warning=FALSE}
ggplot(df, aes(x=Geneset, y=Homo_sapiens.w , fill=Geneset)) +
  geom_violin(width=1.4) +
  geom_boxplot(width=0.1, color="grey", alpha=0.2) +
  labs(y = "dN/dS (ω)") +
  theme_light() +
  theme(legend.position="none", plot.title = element_text(size=11)) + 
  ylim(c(0,1)) +
  stat_summary(
    fun = "mean",
    geom = "point",
    color = "#fff"
  ) +
  stat_summary(
    fun = "mean",
    geom = "text",
    aes(label = round(..y.., 3)),
    hjust = -0.3,  
    color = "#000"
  )
```
